### 2019-03-09 report-gweil.tex
Minor edits.

In the present note, we take as our starting point the naive idea that (1) the sequence of coefficients of a local zeta function and (2) the sequence of coefficients of a global zeta function (or L-function) \emph{should be considered on an equal footing}. From the viewpoint of traditional scheme theory, this doesn't make any sense, but the point is that we might be working towards a new geometric theory which in the local case specializes to the theory of schemes, but which in the global case doesn't even admit a base-change functor to schemes over $\mathrm{Spec} \ \mathbb{Z}$.

@@

\section{Zeta functions and L-functions}

@@

\item When $X$ is a scheme which is flat and of finite type over $\mathrm{Spec} \ \mathbb{Z}$, then the Hasse-Weil zeta function associated to $X$ is defined by
$$ \zeta_X(s) = \prod_{x \in X_{cl}} \left( 1 - N(x)^{-s}   \right)^{-1}  $$
Here $X_{cl}$ is the set of closed points of the scheme $X$, and for any closed point $x$, we have written $N(x)$ for the number of elements in the residue field of $x$. When $X = \mathrm{Spec} \ \mathbb{Z}$ we recover the Riemann zeta function. See \S 6.5 of \cite{Mustata} for more details.
\item For any motive $M$ over $\mathbb{Q}$, we have the motivic L-function $L(M, s)$. Reference: \S 4.5 of \cite{Farmer17}. Examples include Dirichlet L-functions and L-functions of elliptic curves.
\item When $V$ is a variety over a finite field $\mathbb{F}_q$, we have a zeta function
$$  Z_X(t) =   $$

Items 1 and 2 are both examples of \emph{Dirichlet series}, i.e. series of the form $\sum_{n=1}^{\infty} a_n n^-s$, and we refer to them as \emph{global} objects because they incorporate information from \emph{all} primes. Item 3 is an ordinary power series ("MacLaurin series") but can also be written as a Dirichlet series, via the substitution $t \rightsquigarrow q^{-s}$. We write
$$ \zeta_X(s) = Z_X(q^{-s})  $$
for this Dirichlet series. It is of a different nature from the previous ones, since it encodes information from just a single prime (namely the characteristic of the field $\mathbb{F}_q$). We therefore refer to item 3 as a \emph{local} object.

@@

\section{The idea of $m$-functions}

Given a Dirichlet series
$$   \sum_{n = 1}^{\infty}  \frac{a_n}{n^s}   $$
we define the \emph{associated $m$-function} as the power series

$$ \sum_{n = 1}^{\infty} a_n t^{n-1}   $$

We can think of this as just a formal power series (with complex coefficients) but if the Dirichlet series comes from one of the constructions of the previous paragraph, it is in fact a convergent series on the open unit disc, so it defines an analytic function there.

@@

If $L(s)$ is the Dirichlet L-function associated to a Dirichlet character $\chi$, then the $m$-function is a power series which is

\section{ELN triples and the idea of dark schemes}

@@

@@

\section{The global but still rational case (a.k.a. $GL_1$}

@@

>>>>

In the present note, we take as our starting point the incredibly naive idea that (1) the sequence of coefficients of a local zeta function and (2) the sequence of coefficients of a global zeta function (or L-function) \emph{should be considered on an equal footing}. From the viewpoint of traditional scheme theory, this doesn't make any sense, but the point is that we might be working towards a new geometric theory which in the local case specializes to the theory of schemes, but which in the global case doesn't even admit a base-change functor to schemes over $\mathrm{Spec} \ \mathbb{Z}$.

@@

\section{The main background ideas}

\subsection{Zeta functions and L-functions}

@@

\item When $X$ is a scheme which is flat and of finite type over $\mathrm{Spec} \ \mathbb{Z}$, then the Hasse-Weil zeta function associated to $X$ is defined by
$$ \zeta_X(s) = \prod_{x \in X_{cl}} \left( 1 - N(x)^{-s}   \right)^{-1}  $$
Here $X_{cl}$ is the set of closed points of the scheme $X$, and for any closed point $x$, we have written $N(x)$ for the number of elements in the residue field of $x$. When $X = \mathrm{Spec} \ \mathbb{Z}$ we recover the Riemann zeta function. See \S 6.5 of \cite{Mustata} for more details.
\item For any motive $M$ over $\mathbb{Q}$, we have the motivic L-function $L(M, s)$. Reference: \S 4.5 of \cite{Farmer17}. Examples include Dirichlet L-functions and L-functions of elliptic curves. An important remark is that it is perfectly possible to define and study the L-functions of the form $L(h^i(X)(j), s)$ without having to first define what the motive $h^i(X)(j)$ is or in what category it lives.
\item When $V$ is a variety over a finite field $\mathbb{F}_q$, we have a zeta function
$$  Z_X(t) =   $$

Items 1 and 2 are both examples of \emph{Dirichlet series}, i.e. series of the form $\sum_{n=1}^{\infty} a_n n^-s$, and we refer to them as \emph{global} objects because they incorporate information from \emph{all} primes. Item 3 is an ordinary power series ("MacLaurin series") but can also be written as a Dirichlet series, via the substitution $t \rightsquigarrow q^{-s}$. We write
$$ \zeta_X(s) = Z_X(q^{-s})  $$
for this Dirichlet series. It is of a different nature from the previous ones, since it encodes information from just a single prime (namely the characteristic of the field $\mathbb{F}_q$). We therefore refer to item 3 as a \emph{local} object. The coefficients $a_n$ here are \emph{sparse} in the sense that $a_n = 0$ whenever $n$ is not a power of $q$.

@@

\subsection{The idea of $m$-functions}

For any Dirichlet series
$$   \sum_{n = 1}^{\infty}  \frac{a_n}{n^s}   $$
we define \emph{the associated $m$-function} to be

$$ m(t) = a_1 + a_2 t + a_3 t^2 + \ldots = \sum_{n = 1}^{\infty} a_n t^{n-1}   $$

We can think of this as just a formal power series (with complex coefficients) but if the Dirichlet series comes from one of the constructions of the previous paragraph, there are known bounds on the coefficients $a_n$ which imply that the $m$-function is in fact a convergent series on the open unit disc, and hence defines an analytic function there.

@@

If $L(s)$ is the Dirichlet L-function associated to a Dirichlet character $\chi$, then the $m$-function is a power series which is \emph{rational}, since the coefficients form a periodic sequence. A very simple example is given by the unique nontrivial Dirichlet character mod 4, from which we get the $m$-function
$$ 1 - t^2 + \ldots  = \frac{1}{1} $$
(check details).

\subsection{ELN triples}

@@

\subsection{The idea of dark schemes}

\subsection{Functional equations of zzz type}

@@

Explain the meaning of rationality, the zeta function encoding point counts in all finite fields of a given characteristic $p$, which is surprising given that these fields are not related by any maps.

\section{The global but still rational case (a.k.a. $GL_1$)}

@@

\subsection{Weil conjectures over $\mathbb{F}_1$?}

Review here each of the Weil conjectures if needed, and the analogous statements over $\mathbb{F}_1$.

Note that under the analogy we propose, Deligne's Riemann hypothesis in the local setting correspond to global statements that are not at all related to the global Riemann hypothesis, but instead to the BSD conjecture!

\subsection{Further remarks on dark schemes}

We could recall the four viewpoints on schemes (copy from article document section 1.1.).

Mention relation to Tate twist and lack of relation to direct sum and tensor product. We define the differential operator "multiply by t then differentiate", and this corresponds to taking Tate twist on motives.

\subsection{On two-dimensional sequences}

\subsection{Final comments}

Hope for connection to Haran, Toen-Vezzosi, Scholze etc?
